Two applications of logic programming to Coq

### 2020-11-01 papers/types2020/app.tex
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

In Fig.~\ref{fig:interp} we the encoding of the vanilla meta-interpreter used in
the testing phase of the \lsti{dep_pbt} tactic. Differently from Fig.~\ref{fig:augmented} we appeal, via quotations, to Coq's defined connectives --- note the use of a green cut to avoid overlapping with the \lsti{Inductive} atomic case and the absence of certificates.

\begin{figure}
% \newcommand{\XXi}{\Pscr}
% \newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
% \[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
% \infer{\XXi\vdash G_1\wedge G_2}
% {\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
% \]
% \vskip -18pt
% \[
% \infer{\XXi\vdash A}
%       {\bc D A \quad D \in \Pscr}\qquad
%   \infer{\bc A A}{}{}\qquad
%   \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
%   \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
% \]
% In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
% formulas.
 \lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{Encoding in coq-elpi of the vanilla meta-interpreter}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}

@@

\begin{lstlisting}
Inductive has_type : tm -> typ -> Prop :=
   | T_Tru : has_type ttrue TBool
   | T_Fls : has_type tfalse TBool
    | T_Test : forall t1 t2 t3 T,
        has_type t1 TBool -> has_type t2 T -> has_type t3 T ->  has_type (tif  t1 t2 t3) T 
   | T_Zro : has_type tzero TNat
   | T_Scc : forall t1,
         has_type t1 TNat -> has_type (tsucc t1) TNat
   | T_Prd : forall t1,
        has_type t1 TNat -> has_type (tpred t1 ) TNat
   | T_Iszro : forall t1,
        has_type t1 TNat ->  has_type (tiszero t1) TBool.
Inductive nvalue : tm -> Prop :=
| nv_zero : nvalue tzero   | nv_succ : forall t, nvalue t -> nvalue (tsucc t).
Inductive bvalue : tm -> Prop :=
| bv_t : bvalue ttrue   | bv_f : bvalue tfalse. 
 Reserved Notation "t1 '===>' t2" (at level 40).
 Inductive step : tm -> tm -> Prop :=
   | ST_IfTrue : forall t1 t2,
       (tif ttrue t1 t2) ===> t1
   | ST_IfFalse : forall t1 t2,
       (tif tfalse t1 t2) ===> t2
   | ST_If : forall t1 t1' t2 t3,
       t1 ===> t1' -> (tif t1 t2 t3) ===> (tif t1' t2 t3)
   | ST_Succ : forall t1 t1',
       t1 ===> t1' -> (tsucc t1) ===> (tsucc t1')
   | ST_PredZero : (tpred tzero) ===> tzero
   | ST_PredSucc : forall t1,
       nvalue t1 -> (tpred (tsucc t1)) ===> t1
   | ST_Pred : forall t1 t1',
       t1 ===> t1' -> (tpred t1) ===> (tpred t1')
   | ST_IszeroZero : (tiszero tzero) ===> ttrue
   | ST_IszeroSucc : forall t1,
        nvalue t1 ->  (tiszero (tsucc t1)) ===> tfalse
   | ST_Iszero : forall t1 t1',
       t1 ===> t1' -> (tiszero t1) ===> (tiszero t1')
 where "t1 '===>' t2" := (step t1 t2).
Inductive notstuck (e : tm) (Step : tm -> tm -> Prop) : Prop :=
  | pn : nvalue e  -> notstuck e Step
  | pb : bvalue e -> notstuck e Step
  | ps e' : Step e e' -> notstuck e Step.
\end{lstlisting}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

In Fig.~\ref{fig:interp} we report the encoding of the vanilla meta-interpreter used in
the testing phase of the \lsti{dep_pbt} tactic. Differently from Fig.~\ref{fig:augmented} we appeal, via quotations, to Coq's defined connectives. An \lsti{atomic} proposition is one defined \lsti{Inductive}ly. The predicate \lsti{definition} hides the API call to the constructors of such a type.

\begin{figure}
% \newcommand{\XXi}{\Pscr}
% \newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
% \[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
% \infer{\XXi\vdash G_1\wedge G_2}
% {\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
% \]
% \vskip -18pt
% \[
% \infer{\XXi\vdash A}
%       {\bc D A \quad D \in \Pscr}\qquad
%   \infer{\bc A A}{}{}\qquad
%   \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
%   \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
% \]
% In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
% formulas.
 \lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={interp-end}]{code/dep-kernel-v2.mod}
\caption{Encoding in coq-elpi of the vanilla meta-interpreter}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}

@@

We list the rules for static and dynamic semantics of the language mentioned in section~\ref{ssec:dep_pbt} and related notions:
\begin{lstlisting}
Inductive has_type : tm -> typ -> Prop :=
   | T_Tru : has_type ttrue TBool
   | T_Fls : has_type tfalse TBool
    | T_Test : forall t1 t2 t3 T,
        has_type t1 TBool -> has_type t2 T -> has_type t3 T ->  has_type (tif  t1 t2 t3) T 
   | T_Zro : has_type tzero TNat
   | T_Scc : forall t1,
         has_type t1 TNat -> has_type (tsucc t1) TNat
   | T_Prd : forall t1,
        has_type t1 TNat -> has_type (tpred t1 ) TNat
   | T_Iszro : forall t1,
        has_type t1 TNat ->  has_type (tiszero t1) TBool.
Inductive nvalue : tm -> Prop :=
| nv_zero : nvalue tzero   | nv_succ : forall t, nvalue t -> nvalue (tsucc t).
Inductive bvalue : tm -> Prop :=
| bv_t : bvalue ttrue   | bv_f : bvalue tfalse. 
 Reserved Notation "t1 '===>' t2" (at level 40).
 Inductive step : tm -> tm -> Prop :=
   | ST_IfTrue : forall t1 t2,
       (tif ttrue t1 t2) ===> t1
   | ST_IfFalse : forall t1 t2,
       (tif tfalse t1 t2) ===> t2
   | ST_If : forall t1 t1' t2 t3,
       t1 ===> t1' -> (tif t1 t2 t3) ===> (tif t1' t2 t3)
   | ST_Succ : forall t1 t1',
       t1 ===> t1' -> (tsucc t1) ===> (tsucc t1')
   | ST_PredZero : (tpred tzero) ===> tzero
   | ST_PredSucc : forall t1,
       nvalue t1 -> (tpred (tsucc t1)) ===> t1
   | ST_Pred : forall t1 t1',
       t1 ===> t1' -> (tpred t1) ===> (tpred t1')
   | ST_IszeroZero : (tiszero tzero) ===> ttrue
   | ST_IszeroSucc : forall t1,
        nvalue t1 ->  (tiszero (tsucc t1)) ===> tfalse
   | ST_Iszero : forall t1 t1',
       t1 ===> t1' -> (tiszero t1) ===> (tiszero t1')
 where "t1 '===>' t2" := (step t1 t2).
Inductive notstuck (e : tm) (Step : tm -> tm -> Prop) : Prop :=
  | pn : nvalue e  -> notstuck e Step
  | pb : bvalue e -> notstuck e Step
  | ps e' : Step e e' -> notstuck e Step.
\end{lstlisting}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-11-01 papers/types2020/conc.tex
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

% \begin{itemize}
% \item Brief discussion/comparison with \texttt{auto}
% \item Mention other papers by Enrico as applications of coq-elpi~\cite{coen19mscs,tassi19itp}\dots
% \item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in MetaCoq~\cite{sozeau2020metacoq} would be a disaster (reimplement unfication, backtracking, binders)
% \end{itemize}

@@

\todo{Link to the repo}
Extending our tactic to handle computations (\texttt{simp}) and/or to return the partial proof state, unsolved goals in particular. Mention \emph{Harpoon}, Beluga's new tactic language?

>>>>

% \begin{itemize}
% \item Brief discussion/comparison with \texttt{auto}
% \item Mention other papers by Enrico as applications of coq-elpi~\cite{coen19mscs,tassi19itp}\dots [done]
% \item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in MetaCoq~\cite{sozeau2020metacoq} would be a disaster (reimplement unfication, backtracking, binders) [done]
% \end{itemize}

@@

\todo{Link to the repo? \url{https://github.com/proofcert/fpc-elpi}}

Extending our tactic to handle computations (\texttt{simp}) and/or to
return the partial proof state, unsolved goals in particular. Mention
\emph{Harpoon}, Beluga's new tactic language?

### 2020-11-01 papers/types2020/intro.tex
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as a plugin to the Coq
proof assistant \cite{tassi18coqpl}.  This implementation of \lP
extends earlier implementations in primarily two directions:  First,
Elpi adds a notion of constraints and constraints handling rules, which makes it more expressive than
the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
plugin version of Elpi comes equipped with an API for accessing the
Coq environment and a quotation and anti-quotation syntax for mixing
for Coq expressions with \lP program elements.

@@

Since relations (not functions) play a central role in \lP, Elpi is
capable to providing direct support for the many relations that arise
in proof assistants.  Such relations include typing (e.g.,
$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics specifications
\cite{kahn87stacs,hannan93jfp}), and interaction (e.g., structured
operational semantics \cite{plotkin81,milner89book}).  Felty has also
made the point that LCF-style tactics and tacticals can be given an
elegant and natural specification using the higher-order relational
specifications provided by \lP \cite{felty93jar}.  Some recent
implementations built using Elpi
support the usefulness of
higher-order logic programming as a meta-programming language 
for proof assistants in general~\cite{DunchevCT16,coen19mscs} and in particular
for Coq via the \texttt{coq-elpi} plug in~\cite{tassi19itp,CohenST20} .

@@

Before we can present these examples, we first highlight the rather
striking differences in notions of computing and reasoning that arise
on each side of the Elpi-Coq API.   We will also present a quick
summary of the key proof theory concepts that are used by our example
applications.

@@

% LocalWords:  Sacerdoti Coen Tassi Elpi Teyjus Felty LCF tacticals
% LocalWords:  intuitionistic elpi

>>>>

Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as the \texttt{coq-elpi} plugin to the Coq
proof assistant \cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
extends earlier implementations in primarily two directions:  First,
Elpi adds a notion of constraints and constraints handling rules, which makes it more expressive than
the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
plugin version of Elpi comes equipped with  a quotation and anti-quotation syntax for mixing
Coq expressions with \lP program elements and an API for accessing the
Coq environment, typechecking included.

@@

Since relations (not functions) play a central role in \lP, Elpi is
capable to providing direct support for the many relations that have a
role in proof assistants implementation and usage.  Such relations include typing (e.g.,
$\Gamma\vdash M\colon\sigma$), evaluation (e.g., natural semantics
specifications \cite{kahn87stacs,hannan93jfp}), and interaction (e.g.,
structured operational semantics \cite{plotkin81,milner89book}).

Felty has also
made the point that LCF-style tactics and tacticals can be given an
elegant and natural specification using the higher-order relational
specifications provided by \lP \cite{felty93jar}.  Some recent
implementations built using Elpi
support the usefulness of
higher-order logic programming as a meta-programming language 
for proof assistants in general~\cite{DunchevCT16,coen19mscs} and in particular
for Coq via the \texttt{coq-elpi} plug in~\cite{tassi19itp,CohenST20} .
%

@@

While other meta-programming frameworks based on functional
programming such as \emph{MetaCoq}~\cite{sozeau2020metacoq} can and have been used for related endeavors, we believe (together with~\cite{DunchevCT16}) that they would require much more boilerplate.

Before we can present these examples, we first highlight the rather
striking differences in notions of computing and reasoning that arise
on each side of the \texttt{coq-elpi} API.   We will also present a quick
summary of the key proof theory concepts that are used by our example
applications.

@@

% LocalWords:  Sacerdoti Coen Tassi Elpi Teyjus Felty LCF tacticals
% LocalWords:  intuitionistic elpi coq MetaCoq

### 2020-11-01 papers/types2020/pbt.tex
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

\subsection{An Elpi tactic for PBT}
We will invoke the tactic in a proof environment where the overall
goal is % one is trying to prove is the
the property that the system-under-test should meet. This means that,
after \lsti|intro| has been used to introduce the relevant hypotheses,
we have a proof
context % consisting of typing information for a bunch of variables
where eigenvariables play different roles: dependent variables, i.e., the ones appearing also in the types of other variables, are
subject to \emph{generation}, while non-dependent ones tag assumptions that
are only to tested. This is reminiscent of bi-directional
type-checking. % Obviously, the
% two are the result of and \lsti|intro| on either a dependent or a
% non-dependent product
% ; in our setting this division acquires an additional reading beyond
% what the property looks like: since specifications are encoded as pure
% Horn clauses where no dependency is involved, non-dependent variables
% simply tag the specification clauses, whereas dependent variables
% represent the typing information for the data variables that the
% specifications use.

>>>>

\subsection{An Elpi tactic for PBT}
\label{ssec:dep_pbt}
We will invoke the tactic in a proof environment where the overall
goal is % one is trying to prove is the
the property that the system-under-test should meet. This means that,
after \lsti|intro| has been used to introduce the relevant hypotheses,
we have a proof
context % consisting of typing information for a bunch of variables
where eigenvariables play different roles: dependent variables, i.e., the ones appearing also in the types of other variables, are
subject to \emph{generation}, while non-dependent ones tag assumptions that
are only to tested. This is reminiscent of bi-directional
type-checking. % Obviously, the
% two are the result of and \lsti|intro| on either a dependent or a
% non-dependent product
% ; in our setting this division acquires an additional reading beyond
% what the property looks like: since specifications are encoded as pure
% Horn clauses where no dependency is involved, non-dependent variables
% simply tag the specification clauses, whereas dependent variables
% represent the typing information for the data variables that the
% specifications use.

### 2020-11-01 papers/types2020/sec3.tex
sec 4 ready some work on conclusion
related work bits spear thru the paper
ported the appendix and sec3 to kernel-v2

sec3 is still a mess

\begin{metanote}
  (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
 One way to proceed:
  % \begin{itemize}
  % \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
  % \item Then the logic behind the dependent kernel with certificates and proof terms
  % \end{itemize}

Stuff in red is from old paper
\end{metanote}

%\subsection{Horn logic in coq-elpi}

\subsection{Focused proof systems with certificates}
\label{ssec:focused}
\todo{Adapt the discussion}
\todo{grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}

@@

\begin{footnotesize}
\begin{figure}
    \newcommand{\XXi}{{\color{blue}{\Xi}}}

  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi;\Gamma\vdash t : A}
      {\bc{\XXi}{\Gamma} t D A \quad D \in \Pscr\cup\Gamma}
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

@@

\subsection{Certificates}

\todo{Shorten it}
% \todo{Show the dprolog tactic? have an example? could  be
% \lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
{\color{red}
As we mentioned in the introduction, our notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the six expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} (for the
encoding \lsti{B} of a goal formula) is provable in \lP using the
clauses in Figures~\ref{fig:kernel} and~\ref{fig:resources} if and
only if the height of that proof is 5 or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
is provable if and only if the total number of unfoldings of that
proof is 5 or less.

@@

% Finally, Figure~\ref{fig:max} contains the FPC based on the constructor
% \lsti{max} that is used to record explicitly all information within a
% proof: in particular, all disjunctive choices and all substitution
% instances for existential quantifiers are collected into a binary tree
% structure of type \lsti{max}.
% %
% In this sense, proof certificates built with the \lsti{max}
% constructor are \emph{maximally} explicit.
% %
% Such proof certificates are used, for example, in~\cite{Pair}; it is
% important to note that proof checking with such maximally explicit
% certificates can be done with much simpler proof-checkers than those
% used in logic programming since backtracking search and unification
% are not needed.
%

% \begin{figure}
% \lstinputlisting[linerange={max-end}]{code/fpcs.sig}
% \lstinputlisting[linerange={max-end}]{code/fpcs.mod}
% \caption{The \lsti{max} FPC}
% \label{fig:max}
% \end{figure}

Further, if we view a particular FPC as a means of restricting proofs, it is
possible to build an FPC that \emph{restricts} proofs satisfying two FPCs
simultaneously.
%
% In particular, Figure~\ref{fig:pairing} defines an FPC based on the
% (infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
% \lsti{cert}.
%
The pairing experts for the certificate \lsti{Cert1 <c> Cert2} simply
request that the corresponding experts also succeed for both
\lsti{Cert1} and \lsti{Cert2} and, in the case of the \lsti{orE} and
\lsti{someE}, also return the same choice and substitution term,
respectively.
%
Thus, the query 
\begin{lstlisting}
?- check ((qheight 4) <c> (qsize 10)) B
\end{lstlisting}
will succeed if there is a proof of \lsti{B} that has a height less
than or equal to 4 while also being of size less than or equal to 10.
%
% A related use of the pairing of two proof certificates is to %use it to
% \emph{distill} or \emph{elaborate} proof certificates. 
% %
% For example, the proof certificate \lsti{(qsize 5 0)} is rather implicit
% since it will match any proof that used unfold exactly 5 times.
% %
% However, the query
% \begin{lstlisting}
% ?- check ((qsize 5 0) <c> (max Max)) B.
% \end{lstlisting}
% will store into the \lP variable \lsti{Max} more complete details of
% any proof that satisfies the \lsti{(qsize 5 0)} constraint.
% %
% In particular, this forms the infrastructure of 
%  an \emph{explanation} tool for attributing ``blame'' for
% the origin of a counterexample; these maximal
% certificates are an appropriate starting point for documenting
% both the counterexample and why it serves as a counterexample.

% \begin{figure}
% \lstinputlisting[linerange={pairing-end}]{code/fpcs.sig}
% \lstinputlisting[linerange={pairing-end}]{code/fpcs.mod}
% \caption{FPC for pairing}
% \label{fig:pairing}
% \end{figure}

Various  additional examples and experiments using the pairing of FPCs
can be found in~\cite{blanco17cade}. Using similar techniques, it is possible to define FPCs that target
specific types for special treatment: for example, when generating
integers, only (user-defined) small integers could be produced.
}

\subsection{A prolog tactic}

@@

%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE

>>>>

\begin{metanote}
  (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). % In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
% One way to proceed:
  % \begin{itemize}
  % \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
  % \item Then the logic behind the dependent kernel with certificates and proof terms
  % \end{itemize}
%
Stuff in red is from old paper
\end{metanote}

\subsection{Focused proof systems with certificates}
\label{ssec:focused}
\todo{Adapt the discussion}
\todo{grammar for formulae and proof terms \cite{Herbelin94,LengrandDM06}  or just the figure? Mention spine calculus}

@@

\todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
\begin{footnotesize}
\begin{figure}
    \newcommand{\XXi}{{\color{blue}{\Xi}}}

  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi;\Gamma\vdash t : A}
      {\bc{\XXi}{\Gamma} t D A \quad D \in \Pscr\cup\Gamma}
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

@@

%\subsection{Certificates}

\todo{What happened to the unfold certificate???}
% \todo{Show the dprolog tactic? have an example? could  be
% \lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
{\color{red}}
The notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

@@

Further, if we view a particular FPC as a means of restricting proofs, it is
possible to build an FPC that \emph{restricts} proofs satisfying two FPCs
simultaneously.
%
% In particular, Figure~\ref{fig:pairing} defines an FPC based on the
% (infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
% \lsti{cert}.
%
The pairing experts for the certificate \lsti{Cert1 <c> Cert2} simply
request that the corresponding experts also succeed for both
\lsti{Cert1} and \lsti{Cert2} 
%
Thus, the query \lsti{check ((qheight 4) <c> (qsize 10)) B}
will succeed if there is a proof of \lsti{B} that has a height less
than or equal to 4 while also being of size less than or equal to 10.

% Various  additional examples and experiments using the pairing of FPCs
% can be found in~\cite{blanco17cade}. Using similar techniques, it is possible to define FPCs that target
% specific types for special treatment: for example, when generating
% integers, only (user-defined) small integers could be produced.

\subsection{A prolog-like tactic}

@@

The \lsti{dprolog} tactic can be seen as a programmable version of
Coq's \lsti{auto}: it currently lacks the latter capability of solving
goals by \lsti{reflexivity} and the modularity capability of using
\lsti{Hints}, but it is not compelled to use depth-first search, since it follows the dictate of the given FPC. For example the FPC can provide a \emph{trace}, which may be more customable that the one offered by the \lsti{Debug} option.
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity
